Gusto: a toolkit for compatible finite element dynamical cores

### 2024-03-07 source/paper.tex
some progress

\affil[1]{Department of Mathematics and Statistics, University of Exeter}
\affil[2]{Met Office}
\affil[*]{Correspondence to: \texttt{j.shipton@exeter.ac.uk}}

@@

The finite element method represents fields on a mesh made of
non-overlapping, conforming cells, or elements. Within each element,
the field is represented as a sum of coefficients multiplied by
polynomial basis functions. The finite element space is specified by
the degree of the polynomial basis functions and the continuity of
these functions between the elements. The partial differential
equations used in weather and climate modelling have multiple
prognostic fields and it is not necessary, or indeed desirable, to use
the same finite element function space for each field. Using different
finite element function spaces for each field gives what is known as a
mixed finite element discretisation. If the finite element function
spaces are chosen so that

@@

Comparing the equation sets given in the previous sections, we see
several similarities. For example, the prognostic equation for the
velocity always contains a nonlinear transport term, a term due to
rotation and a pressure gradient term. The other prognostic fields
satisfy either an advective or continuity form of the transport
equation. These similarities remain when we formulate the compatible
finite element discretisations, as we shall see below.

@@

@@

In line with the theory outlined in section \ref{}, we
choose $\MM{u} \in \mathbb{V}_1$ and $D \in \mathbb{V}_2$ with
$\mathbb{V}_1 \subset H(\text{div})$ and $\mathbb{V}_2 \subset L^2$. The
corresponding test functions are denoted by $\MM{w}$ and $\phi$. This gives

@@

\JScomment{Text here depends on what the schematic looks like - this
  paragraph should refer to the schematic and explain all aspects}
Prognostic equation sets are each defined in a python class, making
use of the common finite element forms for the transport and
difffusion terms. The timeloop class contains the definition of the
timestepping procedure and the \texttt{run} method that loops over the
timesteps. It also sets up diagnostic fields and calls the IO as
required. The simplest timeloop class will just apply one time
discretisation method to the equation set but it is common in weather
and climate prediction models to split the different types of terms
and apply a time discretisation separately to each type of term. For
example, the transport and forcing terms might be solved for
separately, such as in **, which enables explicit transport schemes to
be applied to each field. ** something about SIQN structure. Terms
related to the physics are solved for in the appropriate part of the
iteration loop, with physics processes that occur on `fast' timescales
evaluated more frequently and `slower' physics processes evaluated
just once at the end of the timestep. The choice of where to evaluate
the physics processes in the iteration is also based on the
computational cost and the bast way to do this is still an open
research question (**ref).

@@

\section{Form Manipulation Labelling}
\label{sec: FML}
The discussion of the time discretisation so far has been general i.e.
not specific to any particular equation set. This is also true of the
implementation of the timestepping schemes. This enables flexibility
and reproducibility (**elaborate). This functionality is enabled
through form manipulation labelling (FML) which is used to associate a
label to each term in the residual form defined in the equation
class. The timestepping methods are then able to reason about which
manipulations to apply to which forms by applying filters. A simple
example would be a splitting method that evaluates transport terms
separately. The full residual is passed to the transport scheme which
then applies a filter to drop all terms that are not labelled as
transport.

@@

>>>>

\affil[1]{Department of Mathematics and Statistics, University of Exeter}
\affil[2]{Met Office}

@@

The finite element method represents fields on a mesh made of
non-overlapping, conforming cells, or elements. Within each element,
the field is represented as a sum of coefficients multiplied by
polynomial basis functions. The finite element space is specified by
the degree of the polynomial basis functions and the continuity of
these functions between the elements. The partial differential
equations used in weather and climate modelling have multiple
prognostic fields and it is not necessary, or indeed desirable, to use
the same finite element function space for each field. Using different
finite element function spaces for each field gives what is known as a
mixed finite element discretisation. If these different finite element
function spaces are chosen so that discrete versions of the vector
calculus identities ** hold then the discretisation is called
`mimetic' or `compatible'.

@@

Comparing the equation sets given in the previous sections, we see
several similarities. For example, the prognostic equation for the
velocity always contains a nonlinear transport term, a term due to
rotation and a `pressure' gradient term (in shallow water this is the
gradient of the depth but the term plays the same role in the dynamics
and is derived from the pressure gradient term in the full equations
(**cite Zeitlin book)). The other prognostic fields satisfy either an
advective or continuity form of the transport equation. These
similarities remain when we formulate the compatible finite element
discretisations, as we shall see below, and importantly this means
that the discretisation of these terms can be written in a general way
that means it can be applied to any of the equation sets.

@@

In line with the theory outlined in section \ref{}, we always choose
the velocity $\MM{u} \in \mathbb{V}_u$ with $\mathbb{V}_u \subset
H(\text{div})$. The field whose gradient appears in the `pressure'
gradient term is chosen to be in the space $\mathbb{V}_p$ with
$\mathbb{V}_p \subset L_2$. The compressible Boussinesq and
compressible Euler equations have a third prognostic variable, either
buoyancy or potential temperature, which is chosen to be in the space
$\mathbb{V}_\theta$ with $\mathbb{V}_\theta \subset ??$ These choices
are governed by the properties required of discretisation of the
linearised equation sets (as discussed in section ** and other
refs). To solve the nonlinear equation sets we need to introduce
stable and accurate discretisations of the transport terms and the
nonlinear pressure gradient term $c_p\theta\nabla \Pi$ in the
compressible Euler velocity equation. Since the prognostic variables
are discretised using different finite element function spaces
$\mathbb{V}_u$, $\mathbb{V}_p$ and $\mathbb{V}_\theta$, each with
different continuity properties between the elements, the choice of
stable transport discretisation is different for each prognostic
variable.

@@

@@

\JScomment{Text here depends on what the schematic looks like - this
  paragraph should refer to the schematic and explain all aspects}
Prognostic equation sets are each defined in a python class, making
use of the common finite element forms for the transport and
difffusion terms. The timeloop class contains the definition of the
timestepping procedure and the \texttt{run} method that loops over the
timesteps. It also sets up diagnostic fields and calls the IO as
required. The simplest timeloop class will just apply one time
discretisation method to the equation set but it is common in weather
and climate prediction models to split the different types of terms
and apply a time discretisation separately to each type of term. For
example, the transport and forcing terms might be solved for
separately, such as in **, which enables explicit transport schemes to
be applied to each field. ** something about SIQN structure. Terms
related to the physics are solved for in the appropriate part of the
iteration loop, with physics processes that occur on `fast' timescales
evaluated more frequently and `slower' physics processes evaluated
just once at the end of the timestep. The choice of where to evaluate
the physics processes in the iteration is also based on the
computational cost and the best way to do this is still an open
research question (**ref).

@@

\section{Form Manipulation Labelling}
\label{sec: FML}
The discussion of the time discretisation so far has been general i.e.
not specific to any particular equation set. This is also true of the
implementation of the timestepping schemes. This enables flexibility
and reproducibility (**elaborate). This functionality is enabled
through form manipulation labelling (FML) which is used to associate a
label to each term in the residual form defined in the equation
class. The timestepping methods are then able to reason about which
manipulations to apply to which forms by applying filters. A simple
example would be a splitting method that evaluates transport terms and
diffusion terms separately. The full residual is passed to the
transport scheme which then applies a filter to drop all terms that
are not labelled as transport; similarly for diffusion. We illustrate
this procedure in figure \ref{fig:advdiff} where we show the
definition of the residual for the advection diffusion equation for a
scalar variable $q$ with a prescribed transport velocity $\MM{c}$ and
diffusivity $\kappa$

\begin{equation}
  q_t + \MM{c}\cdot\nabla q - \kappa\nabla^2 q = 0,
\end{equation}
with the terms labelled either \texttt{time\_derivative},
\texttt{transport} or \texttt{diffusion}. The timestepping loop then
applies a step of forward Euler to the transport term and backwards
Euler to the diffusive term, i.e. the time discretisation is

\begin{align}
  q^* &= q^n - \Delta t (\MM{c}\cdot\nabla) q^n, \\
  q^{n+1} &= q^* + \Delta t \kappa \nabla^2 q^{n+1}.
\end{align}

@@

To include:
\begin{itemize}
\item terminator toy (Tim)
\item shallow water on different grids and with different HDiv families (Nell and Dan)
\item convergence with different RK schemes / multistage schemes (Alex)
\item vertical slice Boussinesq / Euler (+ moisture?)
\item moist baroclinic wave in channel?? (Alex)
\item spherical gravity wave, Boussinesq (Jemma)
\item spherical baroclinic wave, Euler (Dan)
\end{itemize}
